Weighted Model Counting/Integration from the Perspective of Boolean Algebras

### 2020-03-24 paper.tex
It's about ideals, not filters!

% Contributions:
% 1. A generalisation of WMC adapted to Boolean algebras.
% 2. We show that it is a valid measure.
% 3. We show that WMC assumes independence, i.e., cannot represent all measures.
% 4. We show how the BA can be extended with new 'literals' to represent any measure.
% 5. (Maybe) we establish a lower bound on the number of new literals, if we
% make no assumptions about independence.
% 6. (I guess I could) show how evidence = quotiening by a filter, and how
% quotiening by an ideal works
% 7. (WMI) If all weights are 1, we extend Boolean variables to discrete
% variables, and continuous variables are arbitrary subsets (in the
% sigma-algebra), then the WMI process calculates the probability of any event
% and doesn't require going down to the level of models (assuming a suitable
% probability measure).
% 8. If the weight of a literal (and its negation) is 1, then we don't need to
% look at its models--the probability can be calculated at any level. This
% extends to any number of variables.
% 9. How WMI transforms to our algebras. The boring version and what's wrong
% with it.
% 10. Show how to transfer the measure to a quotient algebra and that the answer
% match with each other and with the original definition of WMC.

@@

\begin{definition}[\cite{givant2008introduction}] \label{def:filter}
  A \emph{(Boolean) filter} is a non-empty subset $F \subseteq \mathbf{B}$ such
  that
  \begin{itemize}
  \item $f \land g \in F$ for all $f, g \in F$,
  \item $f \lor a \in F$ for all $f \in F$ and $a \in \mathbf{B}$.
  \end{itemize}
  For any $p \in \mathbf{B}$, the \emph{principal filter of $p$} is the
  intersection $I$ of all filters that contain $p$. It also has the property
  that $p \le a$ for all $a \in I$.
\end{definition}

\begin{definition}[\cite{sikorski1969boolean}] \label{def:quotient}
  Let $F \subseteq \mathbf{B}$ be a filter. The \emph{quotient} $\mathbf{B}/F$
  is a BA of equivalence classes of elements of $\mathbf{B}$ with respect to the
  equivalence relation
  \[
    a \sim b \qquad \iff \qquad a \to b \in F \quad \text{and} \quad b \to a \in F
  \]
  for all $a, b \in \mathbf{B}$. Elements of $\mathbf{B}/F$ are usually denoted
  by $a/F$ (for some $a \in \mathbf{B}$) with the understanding that if $b \sim
  a$ (for some $b \in \mathbf{B}$), then $b/F = a/F$.
\end{definition}

@@

\begin{lemma}
  Let $\mathbf{B}$ be a BA, and let $F$ be the principal filter of $p \in
  \mathbf{B}$. Then, for any $a, b \in \mathbf{B}$, if $a/F = b/F$, then $a
  \land p = b \land p$.
\end{lemma}
\begin{proof}
  If $a/F = b/F$, then $a \to b \in F$ by \cref{def:quotient}. Then $p \le a \to
  b$ by \cref{def:filter}, and $(a \to b) \land p = p$ by the definition of
  $\le$. Then
  \[
    1 = p \lor \neg p = ((a \to b) \lor \neg p) \land (p \lor \neg p) = \neg a
    \lor b \lor \neg p = \neg (a \land p) \lor b.
  \]
  So
  \[
    \neg b = (\neg(a \land p) \land \neg b) \lor (b \land \neg b) = \neg((a
    \land p) \lor b),
  \]
  and thus $b = (a \land p) \lor b$, and
  \[
    b \land p = (a \land p) \lor (b \land p) = (a \lor b) \land p.
  \]
  Similarly, $a \land p = (a \lor b) \land p$ which proves that $a \land p = b
  \land p$.
\end{proof}

\begin{proposition}[Measures on quotients] \label{def:measures_on_quotients}
  Let $\mathbf{B}$ be a BA with a measure $m : \mathbf{B} \to \mathbb{R}_{\ge
    0}$, and let $F$ be the principal filter of $p \in \mathbf{B}$. Let us
  define $m^*\colon \mathbf{B}/F \to \mathbb{R}_{\ge 0}$ as
  \[
    m^*(a/F) = m(a \land p)
  \]
  for any $a \in \mathbf{B}$. The function $m^*$ is well-defined. Furthermore,
  if $m$ is a WMC measure, then so is $m^*$.
\end{proposition}

@@

% TODO: a section on WMI: 3 results
% TODO: introducing evidence can be done by taking a quotient w.r.t. the ideal
% that contradicts the evidence (equivalently, the filter generated by the
% evidence).

>>>>

% Contributions:
% 1. A generalisation of WMC adapted to Boolean algebras.
% 2. We show that it is a valid measure.
% 3. We show that WMC assumes independence, i.e., cannot represent all measures.
% 4. We show how the BA can be extended with new 'literals' to represent any measure.
% 5. (Maybe) we establish a lower bound on the number of new literals, if we
% make no assumptions about independence.
% 6. show how evidence = quotiening by a filter, and how quotiening by an ideal works
% 7. (WMI) If all weights are 1, we extend Boolean variables to discrete
% variables, and continuous variables are arbitrary subsets (in the
% sigma-algebra), then the WMI process calculates the probability of any event
% and doesn't require going down to the level of models (assuming a suitable
% probability measure).
% 8. If the weight of a literal (and its negation) is 1, then we don't need to
% look at its models--the probability can be calculated at any level. This
% extends to any number of variables.
% 9. How WMI transforms to our algebras. The boring version and what's wrong
% with it.
% 10. Show how to transfer the measure to a quotient algebra and that the answer
% match with each other and with the original definition of WMC.

@@

%\begin{definition}[\cite{givant2008introduction}] \label{def:filter}
%  A \emph{(Boolean) filter} is a non-empty subset $F \subseteq \mathbf{B}$ such
%  that
%  \begin{itemize}
%  \item $f \land g \in F$ for all $f, g \in F$,
%  \item $f \lor a \in F$ for all $f \in F$ and $a \in \mathbf{B}$.
%  \end{itemize}
%  For any $p \in \mathbf{B}$, the \emph{principal filter of $p$} is the
%  intersection $I$ of all filters that contain $p$. It also has the property
%  that $p \le a$ for all $a \in I$.
%\end{definition}
%\begin{definition}[\cite{sikorski1969boolean}] \label{def:quotient}
%  Let $F \subseteq \mathbf{B}$ be a filter. The \emph{quotient} $\mathbf{B}/F$
%  is a BA of equivalence classes of elements of $\mathbf{B}$ with respect to the
%  equivalence relation
%  \[
%    a \sim b \qquad \iff \qquad a \to b \in F \quad \text{and} \quad b \to a \in F
%  \]
%  for all $a, b \in \mathbf{B}$. Elements of $\mathbf{B}/F$ are usually denoted
%  by $a/F$ (for some $a \in \mathbf{B}$) with the understanding that if $b \sim
%  a$ (for some $b \in \mathbf{B}$), then $b/F = a/F$.
%\end{definition}

\begin{definition}[\cite{givant2008introduction}] \label{def:ideal}
  An \emph{ideal} is a non-empty subset $I \subseteq \mathbf{B}$ such that
  \begin{itemize}
  \item $i \lor j \in I$ for all $i, j \in I$;
  \item $i \land a \in I$ for all $i \in I$ and $a \in \mathbf{B}$.
  \end{itemize}
  For any $p \in \mathbf{B}$, the \emph{principal ideal of $p$}---denoted by
  $(p)$---is the smallest ideal that contains $p$. It can also be expressed as
  $(p) = \{ a \in \mathbf{B} \mid a \le p \}$.
\end{definition}

\begin{definition}[\cite{givant2008introduction}]
  Let $I$ be an ideal in $\mathbf{B}$. The \emph{quotient algebra of
    $\mathbf{B}$ modulo the ideal $I$} $\mathbf{B}/I$ is a BA
  of equivalence classes of elements of $\mathbf{B}$ with respect to the
  equivalence relation
  \[
    a \sim b \quad \iff \quad a + b \in I
  \]
  for all $a, b \in \mathbf{B}$. Elements of $\mathbf{B}/I$ are usually denoted
  by $a/I$ (for some $a \in \mathbf{B}$) with the understanding that if $b \sim
  a$ (for some $b \in \mathbf{B}$), then $b/I = a/I$. The three algebraic
  operations on $\mathbf{B}/I$ are defined as
  \begin{align*}
    a/I \land b/I &= a \land b / I, \\
    a/I \lor b/I &= a \lor b / I, \\
    \neg(a/I) &= (\neg a)/I.
  \end{align*}
\end{definition}

@@

\begin{lemma}
  For any measure $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$ and elements $a, b
  \in \mathbf{B}$,
  \[
    m(a \lor b) = m(a) + m(b) - m(a \land b).
  \]
\end{lemma}
\begin{proof}
  By \cref{def:measure},
  \begin{align*}
    m(a) &= m(a \land b) + m(a \land \neg b), \\
    m(b) &= m(a \land b) + m(\neg a \land b), \\
    m(a \lor b) &= m(a \land b) + m(a \land \neg b) + m(\neg a \land b),
  \end{align*}
  so
  \[
    m(a) + m(b) - m(a \land b) = m(a \land b) + m(a \land \neg b) + m(\neg a
    \land b) = m(a \lor b)
  \]
  as required.
\end{proof}

\begin{lemma} \label{lemma:well-defined}
  For any $a, b \in \mathbf{B}$ and any principal ideal $(p)$, if $a/(p) =
  b/(p)$, then $a \lor p = b \lor p$.
\end{lemma}
\begin{proof} % TODO: update
  If $a/F = b/F$, then $a \to b \in F$ by \cref{def:quotient}. Then $p \le a \to
  b$ by \cref{def:filter}, and $(a \to b) \land p = p$ by the definition of
  $\le$. Then
  \[
    1 = p \lor \neg p = ((a \to b) \lor \neg p) \land (p \lor \neg p) = \neg a
    \lor b \lor \neg p = \neg (a \land p) \lor b.
  \]
  So
  \[
    \neg b = (\neg(a \land p) \land \neg b) \lor (b \land \neg b) = \neg((a
    \land p) \lor b),
  \]
  and thus $b = (a \land p) \lor b$, and
  \[
    b \land p = (a \land p) \lor (b \land p) = (a \lor b) \land p.
  \]
  Similarly, $a \land p = (a \lor b) \land p$ which proves that $a \land p = b
  \land p$.
\end{proof}

\begin{proposition}[Measures on quotients] \label{def:measures_on_quotients}
  Let $m\colon \mathbf{B} \to \mathbb{R}_{\ge 0}$ be a measure, and let $(p)$ be
  a principal ideal. Let $m^*\colon \mathbf{B}/(p) \to \mathbb{R}_{\ge 0}$ be
  defined as
  \[
    m^*(a/(p)) = m(a \lor p)
  \]
  for any $a \in \mathbf{B}$. The function $m^*$ is well-defined. Furthermore,
  it is a measure on $\mathbf{B}/(p)$ if and only if $m(p) = 0$. Moreover, if it
  is a measure, then the following properties transfer from $m$ to $m^*$:
  \begin{itemize}
  \item if $m$ is a probability measure, then so is $m^*$;
  \item if $m$ is strictly positive, then so is $m^*$;
  \item if $m$ is a WMC measure, then so is $m^*$.
  \end{itemize}
\end{proposition}
\begin{proof} % TODO: prove this
  \Cref{lemma:well-defined} proves that the function is well-defined.
\end{proof}

@@

% TODO: a section on WMI: 3 results
